---
title: "Analysis 1: DS/AI Education Access Shadows in Kansas"
subtitle: "A multiscale enhanced two-step floating catchment area analysis"
author: "Shakil Rafi, Surya Chandramouli, Safia Malallah"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
    theme: cosmo
    fig-width: 11
    fig-height: 6
execute:
  warning: false
  message: false
  echo: false
---

```{r}
library(targets)
library(tidyverse)
library(sf)
library(gt)

atlas <- tar_read(atlas)
data_audit <- tar_read(data_audit)
accessibility <- tar_read(accessibility_classified)
access_map <- tar_read(access_map)
nearest_provider <- tar_read(nearest_provider)
scenario_summary <- tar_read(scenario_summary)
priority_counties <- tar_read(priority_counties_100km)
validation <- tar_read(validation)
accessibility_plot <- tar_read(accessibility_plot)
access_shadow_plot <- tar_read(access_shadow_plot)
scenario_distribution_plot <- tar_read(scenario_distribution_plot)
```

## Research question

Where do Kansas residents have weak physical and digital access to data-science and artificial-intelligence education, and how sensitive are those access patterns to the assumed geographic catchment?

This analysis estimates county-level physical accessibility using an enhanced two-step floating catchment area framework. It then compares physical accessibility with broadband access to identify counties in a baseline **DS/AI access shadow**: physical and digital access at or below their respective statewide medians.

The analysis is descriptive and cross-sectional. It measures potential access under explicit assumptions; it does not measure enrollment, capacity, travel behavior, or causal effects.

## Reproducible data audit

The source contains `r nrow(atlas)` Kansas counties. `r sum(atlas$has_programs == 1)` counties have a positive program-impact score, and `r sum(atlas$online_impact_score > 0)` counties have positive provider-side online impact.

```{r}
data_audit |>
  gt() |>
  cols_label(
    check = "Check",
    value = "Value",
    interpretation = "Interpretation"
  ) |>
  tab_header(title = "Input-data validation") |>
  opt_row_striping()
```

One broadband-coverage value uses `-1` as a missing-data sentinel and is converted to `NA` during analysis. All 105 values of `low_income_digital_access_rate` exceed 1.0; that variable is retained for provenance but excluded from the access-shadow definition pending validation of its numerator and denominator.

## Accessibility model

For provider county \(j\), the supply-to-demand ratio is

\[
R_j = \frac{S_j}{\sum_k P_k w(d_{kj})},
\]

where \(S_j\) is the county's `total_program_impact_score`, \(P_k\) is county population, and \(w(d_{kj})\) is a distance-decay weight.

County \(i\)'s accessibility is

\[
A_i = \sum_j R_j w(d_{ij}).
\]

The baseline kernel is

\[
w(d;c) =
\begin{cases}
\exp\{-3(d/c)^2\}, & d \leq c,\\
0, & d > c,
\end{cases}
\]

evaluated at catchments of 50, 100, and 150 km. Distances are calculated between county representative points in NAD83 / Conus Albers (EPSG:5070). The reported scale is program-impact units per 100,000 potential residents.

## Physical accessibility

```{r}
accessibility_plot
```

The 50 km scenario is intentionally restrictive and leaves `r sum(accessibility$catchment_km == 50 & accessibility$physical_access_per_100k == 0)` counties with zero modeled access. Larger catchments distribute the same statewide provider supply across broader potential-demand areas.

```{r}
scenario_summary |>
  mutate(
    across(
      c(minimum, first_quartile, median, third_quartile, maximum),
      ~ round(.x, 3)
    ),
    access_shadow_population = scales::comma(access_shadow_population)
  ) |>
  gt() |>
  cols_label(
    catchment_km = "Catchment (km)",
    minimum = "Minimum",
    first_quartile = "Q1",
    median = "Median",
    third_quartile = "Q3",
    maximum = "Maximum",
    counties_with_zero_access = "Zero-access counties",
    access_shadow_counties = "Access-shadow counties",
    access_shadow_population = "Access-shadow population"
  ) |>
  tab_header(title = "Accessibility scenario summary") |>
  opt_row_striping()
```

```{r}
scenario_distribution_plot
```

## Physical and digital access shadows

```{r}
access_shadow_plot
```

The access-shadow classification is deliberately simple and auditable. Within each distance scenario, values at or below the statewide median form the low-access group. This convention correctly classifies zero-access counties when the median physical-access score is zero. Results should be treated as a prioritization screen, not as a definitive designation.

### Priority counties under the 100 km scenario

The 100 km scenario is used as the middle baseline because it lies between a restrictive local catchment and a broad regional catchment.

```{r}
priority_counties |>
  mutate(
    population = scales::comma(population),
    physical_access_per_100k = round(physical_access_per_100k, 3),
    broadband_access_index = round(broadband_access_index, 3),
    poverty_rate = scales::percent(poverty_rate, accuracy = 0.1),
    local_program = if_else(local_program, "Yes", "No")
  ) |>
  gt() |>
  cols_label(
    county = "County",
    population = "Population",
    physical_access_per_100k = "Physical access per 100,000",
    broadband_access_index = "Broadband index",
    poverty_rate = "Poverty rate",
    local_program = "Local program"
  ) |>
  tab_header(
    title = "Counties in the 100 km access shadow",
    subtitle = "Ordered from lower physical accessibility and broadband access"
  ) |>
  opt_row_striping()
```

## Nearest-provider diagnostic

Nearest-provider distance is not the accessibility measure, but it is a useful descriptive diagnostic. These distances are straight-line representative-point distances and must not be described as road travel times.

```{r}
nearest_provider |>
  slice_head(n = 15) |>
  mutate(nearest_provider_distance_km = round(nearest_provider_distance_km, 1)) |>
  gt() |>
  cols_label(
    county_name = "County",
    nearest_provider_county = "Nearest Kansas provider county",
    nearest_provider_distance_km = "Distance (km)"
  ) |>
  tab_header(title = "Fifteen counties farthest from a Kansas provider county") |>
  opt_row_striping()
```

## Computational validation

The E2SFCA construction should conserve statewide program-impact supply: population-weighted county accessibility must sum back to the original provider supply under every scenario.

```{r}
validation |>
  mutate(
    across(
      c(original_supply, distributed_supply, absolute_difference),
      ~ format(.x, digits = 8, scientific = FALSE)
    )
  ) |>
  gt() |>
  cols_label(
    catchment_km = "Catchment (km)",
    original_supply = "Original supply",
    distributed_supply = "Distributed supply",
    absolute_difference = "Absolute difference",
    supply_conserved = "Supply conserved",
    all_access_nonnegative = "All access nonnegative",
    county_rows = "County rows"
  ) |>
  tab_header(title = "Model integrity checks") |>
  opt_row_striping()
```

## Interpretation and limitations

1. `total_program_impact_score` is a transparent proxy for educational supply, not enrollment capacity, available seats, faculty, or completions.
2. County representative-point distance is not travel time. Road-network travel times should replace it in a later version.
3. The baseline includes Kansas providers only. Nearby providers in Colorado, Nebraska, Missouri, and Oklahoma may materially change border-county access.
4. `online_impact_score` is assigned to the provider county and does not reveal where online learners live. It is therefore not folded into physical accessibility.
5. Broadband access is used as a separate access dimension, avoiding the circular reuse of `effective_access_score`, which already combines online impact and broadband access.
6. The 50, 100, and 150 km scenarios and the decay parameter are assumptions. Conclusions should be reported only when they are robust across scenarios.

## Reproduction

From the `spatial-analysis` directory:

```r
targets::tar_make()
```

The raw CSV remains unchanged. Derived tables, maps, and validation outputs are written to `output/` by tracked file targets.

```{r}
sessionInfo()
```
