---
title: "Analysis 2: Kansas DS/AI Expansion-Site Optimization"
subtitle: "A transparent population-reach and equity-priority screen"
author: "Shakil Rafi, Surya Chandramouli, Safia Malallah"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show R code"
    embed-resources: true
    theme: cosmo
    fig-width: 11
    fig-height: 6
execute:
  warning: false
  message: false
  echo: false
---

```{r}
library(targets)
library(tidyverse)
library(gt)

atlas <- tar_read(atlas)
expansion_sites <- tar_read(expansion_sites)
expansion_sites_plot <- tar_read(expansion_sites_plot)
```

## Research question

If Kansas could add one to five new DS/AI learning hubs, which county sequence
would expand reach most effectively, and how does an equity objective change
those choices?

This is a deterministic greedy location-allocation screen. It ranks locations
under explicit assumptions; it does not establish capacity, cost, feasibility,
institutional approval, or a real-world optimal facility plan.

## Candidate sites and baseline coverage

The 17 counties with an observed program form the existing network. The other
88 counties are eligible candidate sites. The model uses the same 100 km
Gaussian distance-decay kernel and decay parameter of 3 used in Analysis 1.

For demand county \(i\), baseline reach is the largest distance weight from any
existing provider:

$$
C_i^{(0)} = \max_{j\in\mathcal E} w_{ij}(100).
$$

Taking the maximum prevents overlapping facilities from double-counting the
same resident. Every hypothetical new hub is treated as an equal-strength reach
opportunity.

## Population and equity objectives

The population strategy gives every resident the same weight. The equity
strategy averages standardized poverty, broadband disadvantage, and lower
bachelor's-or-higher attainment:

$$
N_i = \frac{z(\mathrm{poverty}_i)-z(B_i)-z(\mathrm{bachelor}_i)}{3}.
$$

The need score is rescaled to a priority multiplier between 1 and 3. For a
selected candidate set \(\mathcal S\), the strategy-specific objective is

$$
F_q(\mathcal S) = \sum_i P_i q_i
\max\!\left\{C_i^{(0)},\max_{c\in\mathcal S}w_{ic}(100)\right\}.
$$

At each of five rounds, the algorithm evaluates every remaining candidate and
selects the county with the largest marginal gain. Exact ties are broken
alphabetically. Population and equity sequences are optimized independently.

The reported cumulative equivalent reach is distance-weighted population, not
a literal count of students served. Objective gains are comparable within a
strategy, not across the population and equity strategies.

## Results

```{r}
expansion_sites_plot
```

```{r}
expansion_sites |>
  mutate(
    objective_gain = scales::comma(round(objective_gain)),
    cumulative_equivalent_population_covered = scales::comma(
      round(cumulative_equivalent_population_covered)
    ),
    cumulative_zero_access_population_reached = scales::comma(
      cumulative_zero_access_population_reached
    ),
    selected_county_population = scales::comma(
      selected_county_population
    ),
    selected_county_poverty_rate = scales::percent(
      selected_county_poverty_rate,
      accuracy = 0.1
    ),
    selected_county_broadband_index = round(
      selected_county_broadband_index,
      3
    )
  ) |>
  gt(groupname_col = "strategy") |>
  cols_label(
    site_number = "Order",
    selected_county = "Recommended county",
    objective_gain = "Incremental objective gain",
    cumulative_equivalent_population_covered =
      "Cumulative equivalent reach",
    cumulative_zero_access_population_reached =
      "Previously zero-access population reached",
    selected_county_population = "County population",
    selected_county_poverty_rate = "Poverty",
    selected_county_broadband_index = "Broadband index",
    selected_county_four_year_colleges =
      "Atlas four-year-college aggregate (provisional)"
  ) |>
  tab_header(
    title = "One-to-five-site location sequences",
    subtitle = "Conditional on the stated 100 km objective"
  ) |>
  opt_row_striping()
```

## Limitations

1. The greedy sequence is conditional on the modeled objective and is not an
   exact search over every five-county combination.
2. Generic new hubs have no modeled costs, capacities, facilities, staffing,
   political constraints, or institutional partnerships.
3. Representative-point distance is not road travel time, and providers beyond
   Kansas are omitted.
4. Equity components and equal weights are analyst-defined assumptions.
5. The selected-county college field reproduces the Atlas aggregate and is not
   a verified count of distinct institutions.

## Reproduction

From the `spatial-analysis` directory:

```r
targets::tar_make()
```

```{r}
sessionInfo()
```
